### 2017-07-20 memocode_cr/strategy.tex
MEMOCODE'17 cr init

\begin{enumerate}
\item
Determine the context information of the PO: The context information is the definitions specified in the VDM model that are involved in the obligation. The context information may include types, functions, and operations. Note that if a type is involved, its type invariant should be included, too.
\item
Encode context information: this step encodes the context information determined by the first step, and can be divided into several minor steps.
\begin{enumerate}
\item
For a VDM type in the context information, both the type itself and its invariant have to be encoded.
\item
For pre- and post-conditions of a function/operation specified in the VDM model, the encoding is needed if {\tt pre\_<name>} or {\tt post\_<name>} are used in the PO. For example, PO7 of the {\tt CMDS} module described in Section~\ref{section:proof-obligations} mentioned both precondition and postcondition, so both conditions shall be encoded.
\item
If a PO only considers a part of the specification of a type, function, or an operations, it is only needed to encode the expressions that are directly used in the PO. In other words, if a type, a function, or an operation is not directly mentioned, for example, a variable mentioned with {\tt name:type} style, or a function mentioned with {\tt fun\_name(arguments)} style, there is no need to encode the full specification of the type, the function, or the operation.
\end{enumerate}
\item
Obligation negation and quantifier elimination
\begin{enumerate}
\item
Obligation negation: In model checking with SMT solvers, the assertion to be checked is usually negated, and the result of unsatmeans the proof of the assertion succeeds. For the result of sat which means the assertion can be violated, SMT solvers return a model as the counterexample that shows an example of how the assertion is violated. In this work, we also take negation on the PO to be discharged. However, the elimination of forall quantifier should be considered together when applying the negation to the PO.
\item
Elimination of forall quantifiers: POs of VDM usually have forall quantifiers. To solve the formulas with forall quantifiers in Z3, though Z3 accepts formulas with forall quantifiers, there are limitations applying forall quantifiers in Z3 so that by default we have to eliminate forall quantifiers before encoding. In this work, forall quantifier usually causes difficulties in encoding and solving when complex types are involved. Obligation negation described above may achieve the elimination of troublesome foall quantifiers. On the other hand, obligation negation should be applied carefully not to introduce troublesome forall quantifiers from exists quantifiers.
\end{enumerate}
\item
Encode the predicate of obligation and check for satisfiability.
\end{enumerate}

@@

The above Python code imports the z3 module and create a solver object, then declare an integer {\tt i} with a constraint that {\tt i} is greater and equal to zero. In this case, {\tt i} is the encoding an instance of type {\tt nat} in VDM. Note that we cannot add constraints on types but on instance of types in Z3. This means that if there are several variables of the same VDM type that Z3 does not have, instances of the same number of variables have to be declared in Z3 with constraints for every instance added to the solver object. The steps and strategies of encoding discussed above will be demonstrated with details in Section~\ref{section:case-studies} through the report of case studies.

>>>>

\begin{enumerate}
\item
Determine the context information of the PO: The context information is the definitions specified in the VDM model that are involved in the obligation. The context information may include types, functions, and operations. Note that if a type is involved, its type invariant should be included, too.
\item
Encode context information: this step encodes the context information determined by the first step and can be divided into several minor steps.
\begin{enumerate}
\item
For a VDM type in the context information, both the type itself and its invariant have to be encoded.
\item
For pre- and post-conditions of a function/operation specified in the VDM model, the encoding is needed if {\tt pre\_<name>} or {\tt post\_<name>} are used in the PO. For example, PO7 of the {\tt CMDS} module described in Section~\ref{section:proof-obligations} mentioned both precondition and postcondition, so both conditions shall be encoded.
\item
If a PO only considers a part of the specification of a type, a function, or an operations, it is only needed to encode the expressions that are directly used in the PO. In other words, if a type, a function, or an operation is not directly mentioned, for example, a variable mentioned with {\tt name:type} style, or a function mentioned with {\tt fun\_name(arguments)} style, there is no need to encode the full specification of the type, the function, or the operation.
\end{enumerate}
\item
Obligation negation and quantifier elimination
\begin{enumerate}
\item
Obligation negation: In model checking with SMT solvers, the assertion to be checked is usually negated, and the result of unsatmeans the proof of the assertion succeeds. For the result of sat which means the assertion can be violated, SMT solvers return a model as the counterexample that shows an example of how the assertion is violated. In this work, we also take negation on the PO to be discharged. However, the elimination of forall quantifier should be considered together when applying the negation to the PO.
\item
Elimination of forall quantifiers: POs of VDM usually have forall quantifiers. To solve the formulas with forall quantifiers in Z3, though Z3 accepts formulas with forall quantifiers, there are limitations applying forall quantifiers in Z3 so that by default we have to eliminate forall quantifiers before encoding. In this work, forall quantifier usually causes difficulties in encoding and solving when complex types are involved. Obligation negation described above may achieve the elimination of troublesome forall quantifiers. On the other hand, obligation negation should be applied carefully not to introduce troublesome forall quantifiers from exists quantifiers.
\end{enumerate}
\item
Encode the predicate of obligation and check for satisfiability.
\end{enumerate}

@@

The above Python code imports the z3 module and create a solver object, then declare an integer {\tt i} with a constraint that {\tt i} is greater and equal to zero. In this case, {\tt i} is the encoding an instance of type {\tt nat} in VDM. Note that we cannot add constraints on types but instances of types in Z3. This means that if there are several variables of the same VDM type that Z3 does not have, instances of the same number of variables have to be declared in Z3 with constraints for every instance added to the solver object. The steps and strategies of encoding discussed above will be demonstrated with details in Section~\ref{section:case-studies} through the report of case studies.

### 2017-07-20 memocode_cr/studies.tex
MEMOCODE'17 cr init

The first case is a small VDM-SL model specifies an abstract pacemaker. The model is not a full specification and focuses on part of definitions and operations related to heartbeat signals detected and recorded in a pacemaker. For this model of 58 lines, Overture generated nine POs. Here we report the encoding and solving of obligations PO8 and PO3 below.

@@

{\tt Trace} is of type sequence of {\tt Event}, where {\tt Event} is of quote type with two values {\tt <A>} and {\tt <V>} that stand for the A-pace and V-pace of heartbeat signals. At this point, we have necessary context information needed for discharging PO8 and we may proceed to the next step: encoding context information. For PO3, we need to encode types used in PO8: {\tt nat1} and {\tt Trace}. {\tt nat1} is the natural number that excludes zero, so the encoding of {\tt nat1} is as follows.

@@

In the above encoding, {\tt Trace} is encoded as an array type in Z3 and this array type is by default indexed through the integer domain. An array type can not be a sequence type unless we add constraints regarding to sequences to limit an array to a sequence. The idea of adding constraints is the same as the encoding of {\tt nat1} but more complicate. Fig.~\ref{fig:tr_constraint_case1} shows the encoding of the constraint to {\tt tr}, the instance of {\tt Trace} as an array. Note again that the constraints are applied on {\tt tr}, the instance of {\tt Trace}, not the type {\tt Trace}. As shown in Fig.~\ref{fig:tr_constraint_case1}, we added three constraints:

@@

\begin{itemize}
\item
The first constraint (line 4) says that all indices of {\tt tr} below 1 should be undefined because in VDM-SL, a sequence's index starts from 1.
\item
The second constraint (lines 6-13) says that if an index is defined, i.e., not undefined, all indices lower than it should be all defined.
\item
The third constraint (line 15-22) says tat if an index is undefined, all indices greater than it should be all undefined.
\end{itemize}

@@

\begin{enumerate}
\item
(lines 4-7) If {\tt len\_tr} returns 0, then {\tt tr} is an sequence with no element. That is, {\tt tr[1]} has to be undefined so that all indices of {\tt tr} are then undefined based on previous constraints of {\tt tr}. 
\item
(line 8-12) If {\tt len\_tr} returns an integer greater than 0, then {\tt tr} has defined elements till its length index {\tt len\_tr(tr)} such that the element at and after index {\tt len\_tr(tr)+1} should be undefined. 
\end{enumerate}

@@

Finally, we encoded the negated PO8 as the Z3py code shown in Fig.~\ref{fig:po8_encoded_case1}. The encoding has four constraints: two {\tt nat1} variables (line 1); encoding of $P$ ({\tt i in set (inds (tl tr)}) that is actually indices from 1 to {\tt len tr -1} (line 2); encoding of $Q$ ({\tt (i mod aperi) = (vdel + 1)}) (line 3); encoding of $\neg R$ ({\tt not(i in set (inds tr))}) (line 4). Note that {\tt hd} and {\tt tl tr} are operators of the head and tail of a sequence, and {\tt inds} is the operator of gathering indices of defined elements of a sequence and returns a set of natural numbers. Also note that in the encoding in Fig.~\ref{fig:po8_encoded_case1}, we did not explicitly encode {\tt hd}, {\tt tl}, and {\tt inds} and simplified the results of these operators as constraints related to ranges (upper and lower bound) of {\tt tr}. The encoded PO8 (negated) was checked by the solver object in Z3py for satisfiability. The result was {\tt unsat}, which means that PO8 was proved.

@@

The state {\tt Pacemaker} is specified as a pair of natural numbers {\tt aperiod} and {\tt vdelay}. The postcondition of {\tt FaultyHeart} is not specified with {\tt oldstate} which has suffixed tilde, i.e, {\tt aperiod\textasciitilde} and {\tt vdelay\textasciitilde}. Therefore, PO3 is a simplified postcondition satisfiable obligation saying that there exists a {\tt Trace} {\tt tr} which satisfies the postcondition specified with only {\tt newstate}\footnote{How the {\tt newstate} is calculated from {\tt oldstate} is not specified in {\tt FaltyHeart}.}. The postcondition of {\tt FaultyHeart} calls the function {\tt Periodic} twice with three arguments. {\tt Periodic} shown below is an explicitly specified function and returns a boolean value. Therefore, state {\tt Pacemaker}, {\tt post\_FaultyHeart} (postcondition of {\tt FaultyHeart}), and function {\tt Periodic} are the context information of PO3.

@@

Though the specification of {\tt Periodic} is relatively long, the encoding of {\tt Periodic} is similar to the encoding of PO8. It is easy to observe similar computations on the trace {\tt tr} such as {\tt in set inds tr} and {\tt len tr}, and we can encode them in the same way as encoding of PO8. Note that the type {\tt Trace} is again needed in proving PO3 so that the corresponding encoding in PO8 can be used again as constraints here. In the case of PO3, we did not apply negation to the formula since existence quantifier in the prefix of a formula is preferred in Z3. As a result, we got a result of {\tt sat} along with a model that satisfies PO3. Therefore, PO3 was proved.

@@

PO1 shown in Fig.~\ref{fig:po1_case2} is a legal map application obligation that checks whether maps in the system state are applicable. In PO1, state {\tt Exchange} which is a record type of {\tt status} and {\tt calls} is mentioned. Therefore, the context information of PO1 is the state {\tt Exchange} including its invariant. Fig.~\ref{fig:types_case2} shows the specification of state {\tt Exchange} with corresponding quote types. The state {\tt Exchange} contains two maps: {\tt status} indicates the discrete calling status of users, and is a map from {\tt Subscriber} to {\tt Status}; {\tt calls} models the connection relation of users, i.e., which users are connected, and is a map from {\tt Subscriber} to {\tt Subscriber}. Note that {\tt calls} is specified as {\tt inmap} type which means an one-to-one map. This means if a user is calling another user, both of the users can not have other connections to third party users. Also, {\tt Exchange} has state invariant specifying the relations between connections of users and their calling status. At this point, the context information of PO1 is determined.

@@

The strategy of encoding the types shown in Fig.~\ref{fig:types_case2} is (1) applying user-defined types in Z3 to encode the quote types; (2) applying uninterpreted functions to encode map type. In Fig.~\ref{fig:encode_types_case2}, we demonstrate the encoding of selected types. In Fig.~\ref{fig:encode_types_case2}, the type {\tt Recipient} is defined as an user-defined type with values {\tt WR} and {\tt SR}. Type {\tt Status} is further defined as a user-defined type including {\tt Recipient} and {\tt Initiator}. Though {\tt Status} is actually an union type combines {\tt Initiator} and {\tt Recipient}, and can be considered an enumeration, we encoded {\tt Status} using user-defined type style in Z3py. The reason is that we prefer the subtypes {\tt Initiator} and {\tt Status} can be distinguishable in Z3.

@@

status = Function('status',
                  Subscriber,Status_lift)
calls = Function('calls',
                 Subscriber,Subscriber_lift)
\end{Verbatim}
\end{mdframed}
\vspace{-10pt}
\caption{Encoding of quote and map types in Fig.\ref{fig:types_case2} (partial)}
\label{fig:encode_types_case2}
\end{center}
\vspace{-20pt}
\end{figure}

Because of VDM's loic of partial functions, for the type to be declared as the range of map, i.e., {\tt Subscriber} and {\tt Status}, we defined their lifted types in Z3py to take into consideration the case that the value to a key of a map is undefined. Also, though {\tt Subscriber} is of token type which does not have specific values when declared, we treated token types as quote types with predefined domain of values such as {\tt S1}, {\tt S2}. The two maps {\tt status} and {\tt calls} in the state {\tt Exchange} are encoded as uninterpreted functions and the state invariant is encoded by the encoded (and lifted) types mentioned above.

@@

Recall that {\tt calls} is defined as {\tt inmap}, a one-to-one map. Thus, we had to add  the conditions about {\tt inmap} as context information: for any two different users, either both users calls no one (undefined) or they are calling different users. The latter case also includes the situation that only one of the two users is connecting to a third party user and the other user is calling no one.

@@

PO14 shown in Fig.~\ref{fig:PO14_case2} is an enumeration map injectivity obligation of the operation {\tt Answer} of the system. PO14 has free variables, {\tt r} and {\tt status}, in its formula. These free variables are treated as forall quantifier bounded. There are also intermediate variables in the formula such as {\tt m1}, {\tt m2}, {\tt d3}, and {\tt d4}. Note that the map inverse function {\tt inverse} (line 4) and the range restriction operator `{\tt :>}' (line 1) are applied on {\tt calls} and {\tt status} separately. (Recall that {\tt calls} is a one-to-one map {\tt inmap}.) The context information of PO14 includes the two maps {\tt status} and {\tt calls} but not the state {\tt Exchange} because {\tt Exchange} is not mentioned in PO14. Therefore, only the two maps have to be encoded, and the state invariant is excluded in the encoding. The encoding of the constraints for {\tt calls} of type {\tt inmap} described in solving PO1 can be reused here.

@@

When encoding the formula of PO14, firstly, we negated PO14 and eliminated the forall quantifiers. To encode the negated formula of PO14, similar technique for  encoding {\tt set (dom status)} in PO1 can be applied. For the range restriction function, since the range is a singleton set with element value of {\tt <WR>}, we do not have to encode the set notations and only limit the map's range {\tt status(r)} to {\tt <WR>}. Thus, {\tt r in set (dom (status :> \{<WR>\}))} is encoded as follows.

@@

The result of checking PO14 (negated) ware expected to be {\tt unsat}, but we got {\tt sat} instead. Z3 returned a model that satisfies the negated formula of PO14. After examined the model returned by Z3, we found that {\tt calls} in the returned model, {\tt calls} is a map where users calls themselves. This does not violate the type constraint of {\tt inmap}, but an user is not allowed to call him/her-self in a telephone exchange system. Therefore, the checking of PO14 suggested that we should add a constraint on {\tt calls} to the VDM-SL model. We may add the constraint below as the type invariant of {\tt calls} or add it to the state invariant\footnote{We have added the constraint and checked again, and PO14 was proved successfully.}.

@@

As an overall result of our case studies, the objective of our approach that to discharge POs of a VDM-SL model as many as possible is achieved. In the first case study (Abstracted Pacemaker), we discharged all nine POs; in the second case study, we discharged 16 POs our of total 27 POs (about 60\%). To increase the percentage of discharging POs, we may further apply different encoding, for example, encoding maps using arrays in Z3, to solve remaining POs in the Telephone Exchange case study.

% Usability and Applicability of the Encode Style
The two case studies demonstrated the application of the encode strategy described in Section~\ref{section:encode-strategy}. From our experiences, the cost and complexity of encoding varies from specifications in a VDM-SL model. Usually, the complexity of encoded formulas usually depends on the complexity of VDM-SL types. For example, natural number based types such as {\tt seq of nat} or {\tt map from nat to nat} have less complexity, and one may encode and solve this kind of types with SMT solvers without difficulty. On the contrary, for VDM types such as a map of sequences or a set of maps, the encoding may be difficult and the complexity of the encoded formulas may grow a lot.

The encoding demonstrated in the two case studies also showed some encoding patterns. For example, the encoding of union type and quote type with user-defined types or enumeration type, and the encoding of maps with uninterpreted functions, are useful to encoding POs of other VDM-SL models. Also, the encoding patterns like the constraints on arrays to represent a sequence, set inclusion of indices of a sequence, or set inclusion of domain/range of a map, are also reusable. In our case studies, these encoding patterns are actually applied multiple times.

@@

% Scalability and Efficiency}
Our case studies also showed the efficiency of SMT solvers can do great help in discharging POs. Also, the counterexample (model) generation of SMT solvers provides helpful hints for revision and correction of VDM models. As reported in Section~\ref{section:case2}, encoding and solving PO14 of the telephone exchange VDM-SL model found an inconsistency and the counterexample provided a guidance to dissolving the inconsistency. Finally, the scalability of our approach is not serious because we only encode the formula of a PO and its context information. The size of VDM-SL code to be encoded does not escalate with the size of a VDM-SL model.

>>>>

The first case is a small VDM-SL model specifies an abstract pacemaker. The model is not a full specification and focuses on the part of definitions and operations related to heartbeat signals detected and recorded in a pacemaker. For this model of 58 lines, Overture generated nine POs. Here we report the encoding and solving of obligations PO8 and PO3 below.

@@

{\tt Trace} is of type sequence of {\tt Event}, where {\tt Event} is of quote type with two values {\tt <A>} and {\tt <V>} that stand for the A-pace and V-pace of heartbeat signals. At this point, we have necessary context information needed for discharging PO8, and we may proceed to the next step: encoding context information. For PO3, we need to encode types used in PO8: {\tt nat1} and {\tt Trace}. {\tt nat1} is the natural number that excludes zero, so the encoding of {\tt nat1} is as follows.

@@

In the above encoding, {\tt Trace} is encoded as an array type in Z3 and this array type is by default indexed through the integer domain. An array type can not be a sequence type unless we add constraints regarding sequences to limit an array to a sequence. The idea of adding constraints is the same as the encoding of {\tt nat1} but more complicated. Fig.~\ref{fig:tr_constraint_case1} shows the encoding of the constraint to {\tt tr}, the instance of {\tt Trace} as an array. Note again that the constraints are applied on {\tt tr}, the instance of {\tt Trace}, not the type {\tt Trace}. As shown in Fig.~\ref{fig:tr_constraint_case1}, we added three constraints:

@@

\begin{itemize}
\item
The first constraint (line 4) says that all indices of {\tt tr} less than 1 should be undefined because, in VDM-SL, a sequence's index starts from 1.
\item
The second constraint (lines 6-13) says that if an index is defined, i.e., not undefined, all indices lower than it should be all defined.
\item
The third constraint (line 15-22) says tat if an index is undefined, all indices greater than it should be all undefined.
\end{itemize}

@@

\begin{enumerate}
\item
(lines 4-7) If {\tt len\_tr} returns 0, then {\tt tr} is a sequence with no element. That is, {\tt tr[1]} has to be undefined so that all indices of {\tt tr} are then undefined based on previous constraints of {\tt tr}. 
\item
(line 8-12) If {\tt len\_tr} returns an integer greater than 0, then {\tt tr} has defined elements till its length index {\tt len\_tr(tr)} such that the element at and after index {\tt len\_tr(tr)+1} should be undefined. 
\end{enumerate}

@@

Finally, we encoded the negated PO8 as the Z3py code shown in Fig.~\ref{fig:po8_encoded_case1}. The encoding has four constraints: two {\tt nat1} variables (line 1); encoding of $P$ ({\tt i in set (inds (tl tr)}) that is actually indices from 1 to {\tt len tr -1} (line 2); encoding of $Q$ ({\tt (i mod aperi) = (vdel + 1)}) (line 3); encoding of $\neg R$ ({\tt not(i in set (inds tr))}) (line 4). Note that {\tt hd} and {\tt tl tr} are operators of the head and tail of a sequence, and {\tt inds} is the operator of gathering indices of defined elements of a sequence and returns a set of natural numbers. Also, note that in the encoding in Fig.~\ref{fig:po8_encoded_case1}, we did not explicitly encode {\tt hd}, {\tt tl}, and {\tt inds} and simplified the results of these operators as constraints related to ranges (upper and lower bound) of {\tt tr}. The encoded PO8 (negated) was checked by the solver object in Z3py for satisfiability. The result was {\tt unsat}, which means that PO8 was proved.

@@

The state {\tt Pacemaker} is specified as a pair of natural numbers {\tt aperiod} and {\tt vdelay}. The postcondition of {\tt FaultyHeart} is not specified with {\tt oldstate} which has a suffixed tilde, i.e, {\tt aperiod\textasciitilde} and {\tt vdelay\textasciitilde}. Therefore, PO3 is a simplified postcondition satisfiable obligation saying that there exists a {\tt Trace} {\tt tr} which satisfies the postcondition specified with only {\tt newstate}\footnote{How the {\tt newstate} is calculated from {\tt oldstate} is not specified in {\tt FaltyHeart}.}. The postcondition of {\tt FaultyHeart} calls the function {\tt Periodic} twice with three arguments. {\tt Periodic} shown below is an explicitly specified function and returns a boolean value. Therefore, state {\tt Pacemaker}, {\tt post\_FaultyHeart} (postcondition of {\tt FaultyHeart}), and function {\tt Periodic} are the context information of PO3.

@@

Though the specification of {\tt Periodic} is relatively long, the encoding of {\tt Periodic} is similar to the encoding of PO8. It is easy to observe similar computations on the trace {\tt tr} such as {\tt in set inds tr} and {\tt len tr}, and we can encode them in the same way as the encoding of PO8. Note that the type {\tt Trace} is again needed in proving PO3 so that the corresponding encoding in PO8 can be used again as constraints here. In the case of PO3, we did not apply negation to the formula since existence quantifier in the prefix of a formula is preferred in Z3. As a result, we got a result of {\tt sat} along with a model that satisfies PO3. Therefore, PO3 was proved.

@@

PO1 shown in Fig.~\ref{fig:po1_case2} is a legal map application obligation that checks whether maps in the system state are applicable. In PO1, state {\tt Exchange} which is a record type of {\tt status} and {\tt calls} is mentioned. Therefore, the context information of PO1 is the state {\tt Exchange} including its invariant. Fig.~\ref{fig:types_case2} shows the specification of state {\tt Exchange} with corresponding quote types. The state {\tt Exchange} contains two maps: {\tt status} indicates the discrete calling status of users, and is a map from {\tt Subscriber} to {\tt Status}; {\tt calls} models the connection relation of users, i.e., which users are connected, and is a map from {\tt Subscriber} to {\tt Subscriber}. Note that {\tt calls} is specified as {\tt inmap} type which means a one-to-one map. This means if a user is calling another user, both of the users can not have other connections to third party users. Also, {\tt Exchange} has state invariant specifying the relations between connections of users and their calling status. At this point, the context information of PO1 is determined.

@@

The strategy of encoding the types shown in Fig.~\ref{fig:types_case2} is (1) applying user-defined types in Z3 to encode the quote types; (2) applying uninterpreted functions to encode map type. In Fig.~\ref{fig:encode_types_case2}, we demonstrate the encoding of selected types. In Fig.~\ref{fig:encode_types_case2}, the type {\tt Recipient} is defined as an user-defined type with values {\tt WR} and {\tt SR}. Type {\tt Status} is further defined as a user-defined type including {\tt Recipient} and {\tt Initiator}. Though {\tt Status} is a union type combines {\tt Initiator} and {\tt Recipient}, which can be considered an enumeration, we encoded {\tt Status} using user-defined type style in Z3py. The reason is that we prefer the subtypes {\tt Initiator} and {\tt Status} can be distinguishable in Z3.

@@

status = Function('status',
                  Subscriber,Status_lift)
calls = Function('calls',
                 Subscriber,Subscriber_lift)
\end{Verbatim}
\end{mdframed}
\vspace{-10pt}
\caption{Encoding of quote and map types in Fig.\ref{fig:types_case2} (partial)}
\label{fig:encode_types_case2}
\end{center}
\vspace{-15pt}
\end{figure}

Because of VDM's logic of partial functions, for the type to be declared as the range of the map, i.e., {\tt Subscriber} and {\tt Status}, we defined their lifted types in Z3py to take into consideration the case that the value of a key of a map is undefined. Also, though {\tt Subscriber} is of token type which does not have specific values when declared, we treated token types as quote types with a predefined domain of values such as {\tt S1}, {\tt S2}. The two maps {\tt status} and {\tt calls} in the state {\tt Exchange} are encoded as uninterpreted functions and the state invariant is encoded by the encoded (and lifted) types mentioned above.

@@

Recall that {\tt calls} is defined as {\tt inmap}, a one-to-one map. Thus, we had to add  the conditions about {\tt inmap} as context information: for any two different users, either both users calls no one (undefined) or they are calling different users. Note that the latter case also includes the situation that only one of the two users is connecting to a third party user and the other user is calling no one.

@@

PO14 shown in Fig.~\ref{fig:PO14_case2} is an enumeration map injectivity obligation of the operation {\tt Answer} of the system. PO14 has free variables, {\tt r} and {\tt status}, in its formula. These free variables are treated as forall quantifier bounded. There are also intermediate variables in the formula such as {\tt m1}, {\tt m2}, {\tt d3}, and {\tt d4}. Note that the map inverse function {\tt inverse} (line 4) and the range restriction operator `{\tt :>}' (line 1) are applied to {\tt calls} and {\tt status} separately. (Recall that {\tt calls} is a one-to-one map {\tt inmap}.) The context information of PO14 includes the two maps {\tt status} and {\tt calls} but not the state {\tt Exchange} because {\tt Exchange} is not mentioned in PO14. Therefore, only the two maps have to be encoded, and the state invariant is excluded in the encoding. The encoding of the constraints for {\tt calls} of type {\tt inmap} described in solving PO1 can be reused here.

@@

When encoding the formula of PO14, firstly, we negated PO14 and eliminated the forall quantifiers. To encode the negated formula of PO14, a similar technique for  encoding {\tt set (dom status)} in PO1 can be applied. For the range restriction function, since the range is a singleton set with an element {\tt <WR>}, we do not have to encode the set notations and only limit the map's range {\tt status(r)} to {\tt <WR>}. Thus, {\tt r in set (dom (status :> \{<WR>\}))} is encoded as follows.

@@

The result of checking PO14 (negated) was expected to be {\tt unsat}, but we got {\tt sat} instead. Z3 returned a model that satisfies the negated formula of PO14. After examined the model returned by Z3, we found that {\tt calls} in the returned model, {\tt calls} is a map where users call themselves. This does not violate the type constraint of {\tt inmap}, but a user is not allowed to call him/her-self in a telephone exchange system. Therefore, the checking of PO14 suggested that we should add a constraint on {\tt calls} to the VDM-SL model. We may add the constraint below as the type invariant of {\tt calls} or add it to the state invariant\footnote{We have added the constraint and checked again, and PO14 was proved successfully.}.

@@

As an overall result of our case studies, the objective of our approach that to discharge POs of a VDM-SL model as many as possible is considered achieved. In the first case study (Abstracted Pacemaker), we discharged all nine POs; in the second case study, we discharged 16 POs our of total 27 POs (about 60\%). To increase the percentage of discharging POs, we may further apply different encoding, for example, encoding maps using arrays in Z3, to solve remaining POs in the Telephone Exchange case study.

% Usability and Applicability of the Encode Style
The two case studies demonstrated the application of the encode strategy described in Section~\ref{section:encode-strategy}. From our experiences, the cost and complexity of encoding varies from specifications in a VDM-SL model. Usually, the complexity of encoded formulas usually depends on the complexity of VDM-SL types. For example, natural number based types such as {\tt seq of nat} or {\tt map from nat to nat} have less complexity, and one may encode and solve this kind of types with SMT solvers without difficulty. On the contrary, for VDM types such as a map of sequences or a set of maps, the encoding may be difficult, and the complexity of the encoded formulas may grow a lot.

The encoding demonstrated in the two case studies also showed some encoding patterns. For example, the encoding of union type and quote type with user-defined types or enumeration type, and the encoding of maps with uninterpreted functions, are useful to encoding POs of other VDM-SL models. Also, the encoding patterns like the constraints on arrays to represent a sequence, set inclusion of indices of a sequence, or set inclusion of domain/range of a map, are also reusable. In our case studies, these encoding patterns are applied multiple times.

@@

% Scalability and Efficiency}
Our case studies also showed the efficiency of SMT solvers can do great help in discharging POs. Also, the counterexample (model) generation of SMT solvers provides helpful hints for revision and correction of VDM models. As reported in Section~\ref{section:case2}, encoding and solving PO14 of the telephone exchange VDM-SL model found an inconsistency and the counterexample provided guidance to dissolving the inconsistency. Finally, the scalability of our approach is not serious because we only encode the formula of a PO and its context information. The size of VDM-SL code to be encoded does not escalate with the size of a VDM-SL model.

### 2017-07-20 memocode_cr/vdm.tex
MEMOCODE'17 cr init

% General information of VDM
The Vienna Development Method (VDM) was originally developed in the 1970's at the IBM laboratories in Vienna~\cite{DBLP:conf/fm/1978}. The VDM Specification Language (VDM-SL) is a higher-order language with formally defined syntax and semantics~\cite{Prehn:1991:LNCS551,Larsen1995585}. VDM provides various abstract data types: basic types such as boolean, natural number, and token; advanced types such as record, product, set, and map. A state consists of typed variables. Typed variables in a state may be restricted by invariants and operations/functions may be specified with preconditions and postconditions.

@@

% Explaining the sample VDM-SL model: part2
The operation {\tt push\_cmd} is specified to accept a command {\tt CMD} or {\tt nil} and pushes the command as the head of {\tt commands}. The precondition demands that \texttt{commands} is an empty sequence or its head is different from the argument. The postcondition is specified using the old state, the state before the execution of {\tt push\_cmd}, that denoted as {\tt commands\textasciitilde} with postfixed tilde. The postcondition says that the old state is the tail of the new state {\tt commands}. Observe that {\tt push\_cmd} is an implicit operation that only specifies the signature with pre- and post-conditions. The function \texttt{times\_count} that counts the occurrences of commands is specified explicitly. Instead of using loop, the function uses sequence comprehension to construct the map of type {\tt CMD\_times}. Specifically, the resultant map has the domain $\{ \texttt{<R>}, \texttt{<L>} \}$. It maps $\texttt{<R>}$ to the length of the sequence which contains the indices of $\texttt{a}$ having the value $\texttt{<R>}$. That is, the resultant map gives the number of $\texttt{<R>}$ in the sequence $\texttt{a}$ on $\texttt{<R>}$. Simiarly, it gives the number of $\texttt{<L>}$ in $\texttt{a}$ on $\texttt{<L>}$. This also shows the expression power of VDM in specification.

>>>>

% General information of VDM
The Vienna Development Method (VDM) was originally developed in the 1970's at the IBM laboratories in Vienna~\cite{DBLP:conf/fm/1978}. The VDM Specification Language (VDM-SL) is a higher-order language with formally defined syntax and semantics~\cite{Prehn:1991:LNCS551,Larsen1995585}. VDM provides various abstract data types: basic types such as Boolean, natural number, and token; advanced types such as record, product, set, and map. A state consists of typed variables. Typed variables in a state may be restricted by invariants and operations/functions may be specified with preconditions and postconditions.

@@

% Explaining the sample VDM-SL model: part2
The operation {\tt push\_cmd} is specified to accept a command {\tt CMD} or {\tt nil} and pushes the command as the head of {\tt commands}. The precondition demands that \texttt{commands} is an empty sequence or its head is different from the argument. The postcondition is specified using the old state, the state before the execution of {\tt push\_cmd}, which is denoted as {\tt commands\textasciitilde} with a postfixed tilde. The postcondition says that the old state is the tail of the new state {\tt commands}. Observe that {\tt push\_cmd} is an implicit operation that only specifies the signature with pre- and post-conditions. The function \texttt{times\_count} that counts the occurrences of commands is specified explicitly. Instead of using a loop, the function uses sequence comprehension to construct the map of type {\tt CMD\_times}. Specifically, the resultant map has the domain $\{ \texttt{<R>}, \texttt{<L>} \}$. It maps $\texttt{<R>}$ to the length of the sequence which contains the indices of $\texttt{a}$ having the value $\texttt{<R>}$. That is, the resultant map gives the number of $\texttt{<R>}$ in the sequence $\texttt{a}$ on $\texttt{<R>}$. Similarly, it gives the number of $\texttt{<L>}$ in $\texttt{a}$ on $\texttt{<L>}$. This also shows the expression power of VDM in specification.
